---
title: "FDA Project"
author: "Oumaima El Menni"
date: "2024-11-04"
output: html_document
---

# Loading packages


```{r setup, include=FALSE}

library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)
library(fda)
library(tidyverse)
library(pheatmap)
library(knitr)
library(kableExtra)
library(zoo)       
library(gridExtra)
library(refund)
library(dtw)
library(fdapace)
library(depthTools)
library(rainbow)
library(roahd)
library(ggforce)
library(reshape2)
library(fda.usc)
library(pracma)
library(refund)
library(viridis)

```

# Loading Dataset

```{r cars}
file_path <- "E:\\Bureau\\Study\\Y2\\Functional and Topological Data Analysis\\Data Stocks Crypto.xlsx"

data <- read_excel(file_path)
head(data, 5)

```
```{r pressure, echo=FALSE}
data <- na.omit(data)
data$date <- as.Date(data$date)

head(data, 5)
```

## Visualization


```{r pressure, echo=FALSE}
assets <- c("AAPL", "JNJ", "KO", "WMT", "XOM", "PG", "JPM", "UNH", "MCD", 
            "NKE", "ETH", "DSH", "XLM", "PPC", "DOGE", "XEM", "XRP", "XMR", 
            "BTC", "LTC")


stocks <- c("AAPL", "JNJ", "KO", "WMT", "XOM", "PG", "JPM", "UNH", "MCD", "NKE")
cryptos <- c("ETH", "DSH", "XLM", "PPC", "DOGE", "XEM", "XRP", "XMR", "BTC", "LTC")


data_long <- data %>%
  pivot_longer(cols = all_of(assets),  # Use 'all_of' to refer to the vector
               names_to = "Asset", 
               values_to = "Price")


for (asset_name in assets) {
  

    asset_data <- data_long %>% filter(Asset == asset_name)
  

      plot <- ggplot(asset_data, aes(x = date, y = Price)) +
    geom_line(color = "blue") +
    labs(title = paste("Price of", asset_name, "Over Time"), 
         x = "Date", y = "Price") +
    theme_minimal()
  
  

        print(plot)
} 
```


# Standardization

```{r pressure, echo=FALSE}

for (asset in assets) {
  data[[paste0(asset, "_scaled")]] <- (data[[asset]] - min(data[[asset]])) / 
                                       (max(data[[asset]]) - min(data[[asset]]))
}

```


# B-Spline Smoothing 

```{r pressure, echo=FALSE}
# Defining assets with corresponding nbasis values (chosen based on visual inspection)
assets_nbasis <- c(AAPL = 35, JNJ = 25, KO = 25, WMT = 35, XOM = 50, PG = 30, 
                   JPM = 30, UNH = 18, MCD = 30, NKE = 45, ETH = 30, DSH = 20, 
                   XLM = 25, PPC = 40, DOGE = 30, XEM = 30, XRP = 30, XMR = 35, 
                   BTC = 30, LTC = 40)

basis_smooth_list <- list()  

for (asset in names(assets_nbasis)) {
  nbasis_val <- assets_nbasis[asset]
  
  basis <- create.bspline.basis(rangeval = range(data$date), nbasis = nbasis_val)
  
  smooth_result <- smooth.basis(data$date, data[[paste0(asset, "_scaled")]], basis)
  
  basis_smooth_list[[paste0(asset, "_smooth")]] <- smooth_result
}


```

## Original Curves vs Smoothed Curves

```{r pressure, echo=FALSE}

for (asset in names(assets_nbasis)) {
  
  
  plot(data$date, data[[paste0(asset, "_scaled")]], type = "l", col = "blue", lty = 1,
       main = paste(asset, "Stock Price - B-spline Smoothing"), xlab = "Date", ylab = "Price")
  
  
  lines(basis_smooth_list[[paste0(asset, "_smooth")]]$fd, col = "red", lty = 2)
  
  
  legend("topleft", legend = c(paste(asset, "Original"), paste(asset, "Smoothed")),
         col = c("blue", "red"), lty = 1:2)
  
  
  Sys.sleep(1)  
}

```
# Visualization

```{r pressure, echo=FALSE}

plot(data$date, data$AAPL_scaled, type = "l", col = "blue", 
     ylim = range(data[, paste0(names(assets_nbasis), "_scaled")], na.rm = TRUE), 
     main = "Scaled Prices of All Assets", xlab = "Date", ylab = "Scaled Price")


colors <- viridis(length(assets_nbasis))  
i <- 1
for (asset in names(assets_nbasis)) {
  lines(data$date, data[[paste0(asset, "_scaled")]], col = colors[i], lty = 1)
  i <- i + 1
}

```


```{r pressure, echo=FALSE}


# Plot 1: Original scaled data for all assets
plot(data$date, data$AAPL_scaled, type = "l", col = colors[1],
     ylim = range(data[, paste0(names(assets_nbasis), "_scaled")], na.rm = TRUE),
     xlim = range(data$date),  
     main = "Scaled Prices of All Assets", xlab = "Date", ylab = "Scaled Price")

# Loop to add each asset's scaled data
for (i in seq_along(names(assets_nbasis))) {
  asset <- names(assets_nbasis)[i]
  lines(data$date, data[[paste0(asset, "_scaled")]], col = colors[i], lty = 1)
}


# Plot 2: Smoothed curves for all assets
plot(data$date, rep(NA, length(data$date)),  # Empty plot to set up x-axis range
     ylim = c(-0.1, 1.1), xlim = range(data$date),  # Set y-axis limits as required
     main = "Smoothed Curves of All Assets", xlab = "Date", ylab = "Smoothed Price")

# Loop to add each asset's smoothed data
for (i in seq_along(names(assets_nbasis))) {
  asset <- names(assets_nbasis)[i]
  lines(basis_smooth_list[[paste0(asset, "_smooth")]]$fd, col = colors[i], lty = 1)
}

```



# Creating functional objects

```{r pressure, echo=FALSE}
fd_list <- list()

for (asset in assets) {
  fd_list[[paste0(asset, "_fd")]] <- basis_smooth_list[[paste0(asset, "_smooth")]]$fd
}
```



# Functional Cross-Correlation




```{r pressure, echo=FALSE}

# To store cross-correlation results
cross_cor_results <- list()


for (stock in stocks) {
  for (crypto in cryptos) {
    # Extract functional data objects for the stock and crypto
    stock_fd <- fd_list[[paste0(stock, "_fd")]]
    crypto_fd <- fd_list[[paste0(crypto, "_fd")]]
    

    
    ccf_result <- inprod(stock_fd, crypto_fd)  
    
    
    cross_cor_results[[paste(stock, crypto, sep = "_")]] <- ccf_result
  }
}


cross_cor_summary <- data.frame(Stock = character(), Crypto = character(), Corr = numeric())

for (key in names(cross_cor_results)) {
  stock_crypto <- strsplit(key, "_")[[1]]
  stock <- stock_crypto[1]
  crypto <- stock_crypto[2]
  
  corr <- cross_cor_results[[key]]  # Access the stored inner product result
  
  cross_cor_summary <- rbind(cross_cor_summary, data.frame(Stock = stock, Crypto = crypto, Corr = corr))
}

print(cross_cor_summary)

corr_min <- min(cross_cor_summary$Corr, na.rm = TRUE)
corr_max <- max(cross_cor_summary$Corr, na.rm = TRUE)

heatmap_data <- dcast(cross_cor_summary, Stock ~ Crypto, value.var = "Corr")
ggplot(melt(heatmap_data, id.vars = "Stock"), aes(x = variable, y = Stock, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = (corr_min + corr_max) / 2, limit = c(corr_min, corr_max), 
                       name="Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Heatmap of Cross-Correlation between Stocks and Cryptocurrencies",
       x = "Cryptocurrencies", 
       y = "Stocks")

```
# Rolling Correlation
## Heatmap


```{r pressure, echo=FALSE}
data$Date <- as.Date(data$date, format = "%Y-%m-%d")

time_points <- as.numeric(data$date)
```



#Heatmap of rolling volatility


```{r pressure, echo=FALSE}

time_grid <- data$date  # Extract the date column

calculate_rolling_volatility <- function(fd, time_grid, window_size) {
    
    evaluations <- eval.fd(time_grid, fd)
    
    
    vol <- rollapply(evaluations, width = window_size, FUN = sd, by = 1, fill = NA, align = "right")
    
    return(vol)
}

# Parameters for rolling volatility
window_size <- 10  


rolling_volatility <- list()

for (asset in c(stocks, cryptos)) {
    
    fd_object <- fd_list[[paste0(asset, "_fd")]]
    
    
    rolling_volatility[[paste0(asset, "_vol")]] <- calculate_rolling_volatility(fd_object, time_grid, window_size)
}


volatility_matrix <- do.call(cbind, lapply(c(stocks, cryptos), function(asset) {
    vol <- rolling_volatility[[paste0(asset, "_vol")]]
    ifelse(is.na(vol), 0, vol)  
}))


rownames(volatility_matrix) <- time_grid  
colnames(volatility_matrix) <- c(stocks, cryptos)


volatility_df <- melt(volatility_matrix)

volatility_df$Var1 <- as.Date(volatility_df$Var1)


ggplot(volatility_df, aes(x = Var1, y = Var2, fill = value)) +
    geom_tile() +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = median(volatility_df$value, na.rm = TRUE)) +
    labs(title = "Heatmap of Rolling Volatility", x = "Date", y = "Assets", fill = "Volatility") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank()) +
    scale_x_date(limits = c(min(volatility_df$Var1), max(volatility_df$Var1)))  # Set x-axis limits

```


# fANOVA: Compare average behavior 

```{r pressure, echo=FALSE}
fd_stocks <- lapply(stocks, function(asset) fd_list[[paste0(asset, "_fd")]])
fd_cryptos <- lapply(cryptos, function(asset) fd_list[[paste0(asset, "_fd")]])

min_length <- min(sapply(c(fd_stocks, fd_cryptos), function(x) length(x$coefs)))


fd_stocks <- lapply(fd_stocks, function(fd) fd$coefs[1:min_length, ])
fd_cryptos <- lapply(fd_cryptos, function(fd) fd$coefs[1:min_length, ])


mean_stocks <- rowMeans(do.call(cbind, fd_stocks), na.rm = TRUE)
mean_cryptos <- rowMeans(do.call(cbind, fd_cryptos), na.rm = TRUE)

mean_combined <- rowMeans(cbind(mean_stocks, mean_cryptos), na.rm = TRUE)


n_stocks <- length(stocks)
n_cryptos <- length(cryptos)


SSB <- n_stocks * sum((mean_stocks - mean_combined)^2) + 
       n_cryptos * sum((mean_cryptos - mean_combined)^2)

# Calculate SSW 
SSW_stocks <- sum(sapply(fd_stocks, function(fd) sum((fd - mean_stocks)^2, na.rm = TRUE)))
SSW_cryptos <- sum(sapply(fd_cryptos, function(fd) sum((fd - mean_cryptos)^2, na.rm = TRUE)))

SSW <- SSW_stocks + SSW_cryptos

# Calculate degrees of freedom
df_between <- 1  # 2 groups - 1
df_within <- (n_stocks + n_cryptos) - 2  

# Calculate the mean squares
MSB <- SSB / df_between
MSW <- SSW / df_within

# Calculate the F-statistic
F_statistic <- MSB / MSW


cat("F-statistic:", F_statistic, "\n")
cat("Degrees of freedom (between):", df_between, "\n")
cat("Degrees of freedom (within):", df_within, "\n")

# Calculate the p-value based on F-distribution
p_value <- pf(F_statistic, df_between, df_within, lower.tail = FALSE)
cat("p-value:", p_value, "\n")

```
## Mean Functions of Stocks and Cryptos

```{r pressure, echo=FALSE}



time_points <- seq(1, min_length)  # Adjust according to your actual time points
mean_data <- data.frame(
  Time = rep(time_points, 2),
  MeanValue = c(mean_stocks, mean_cryptos),
  Group = rep(c("Stocks", "Cryptos"), each = min_length)
)

# Plot
ggplot(mean_data, aes(x = Time, y = MeanValue, color = Group)) +
  geom_line() +
  labs(title = "Mean Functions of Stocks and Cryptos",
       x = "Time",
       y = "Mean Value") +
  theme_minimal()

```


# Dynamic Time warping


```{r pressure, echo=FALSE}
# Define the DTW matrix to store distances
dtw_matrix <- matrix(NA, nrow = length(stocks), ncol = length(cryptos),
                     dimnames = list(stocks, cryptos))

# Define the evaluation points as the dates in the dataset
eval_points <- data$date

# Function to calculate DTW distances in parallel
cl <- makeCluster(detectCores() - 1)  # Initialize the cluster (using all but one core)

# Export necessary functions and variables to the parallel workers
clusterExport(cl, list("stocks", "cryptos", "fd_list", "dtw", "eval_points"))

# Export the 'fda' package and any other required libraries to the cluster
clusterEvalQ(cl, library(fda))

# Calculate DTW distances between each stock and cryptocurrency in parallel
dtw_matrix <- parSapply(cl, stocks, function(stock) {
  sapply(cryptos, function(crypto) {
    # Evaluate the functional data for the stock and cryptocurrency
    stock_fd_values <- eval.fd(eval_points, fd_list[[paste0(stock, "_fd")]])
    crypto_fd_values <- eval.fd(eval_points, fd_list[[paste0(crypto, "_fd")]])
    
    # Perform DTW between the evaluated time series
    dtw_result <- dtw(stock_fd_values, crypto_fd_values)
    
    # Return the DTW distance
    return(dtw_result$distance)
  })
})

# Stop the cluster after the computation
stopCluster(cl)

# Convert the DTW matrix to a data frame for visualization
dtw_df <- as.data.frame(as.table(dtw_matrix))
colnames(dtw_df) <- c("Stock", "Crypto", "DTW_Distance")

# Plot a heatmap to visualize DTW distances between stocks and cryptos
ggplot(dtw_df, aes(x = Stock, y = Crypto, fill = DTW_Distance)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "blue", high = "red") +
  geom_text(aes(label = round(DTW_Distance, 2)), color = "white", size = 3) +  
  labs(title = "DTW Distances Between Stocks and Cryptocurrencies",
       x = "Stock", y = "Cryptocurrency", fill = "DTW Distance") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```





# Depth Analysis

```{r pressure, echo=FALSE}

day_range <- c(1, nrow(data))  
common_basis <- create.bspline.basis(day_range, nbasis=15)


smooth_data_list <- lapply(c(stocks, cryptos), function(asset) {
  smooth.basis(argvals=1:nrow(data), y=data[[paste0(asset, "_scaled")]], fdParobj=common_basis)
})

coefs_matrix <- do.call(cbind, lapply(smooth_data_list, function(x) x$fd$coefs))

# Create the combined fd object for all assets
combined_fd <- fd(coef=coefs_matrix, basisobj=common_basis, 
                  fdnames=list("Time", c(stocks, cryptos), "Scaled Price"))

colors <- c(rep("red", length(stocks)), rep("black", length(cryptos)))

date_values <- data$date

plot(combined_fd, col=colors, lty=1, 
     xlab="Date", ylab="Scaled Price", main="Depth Analysis of Stocks and Cryptos",
     xaxt = "n")  # Suppress default x-axis

tick_positions <- seq(1, nrow(data), length.out = 3)  # 10 ticks for better visibility
tick_labels <- format(date_values[tick_positions], "%Y")
axis(1, at=tick_positions, labels=tick_labels, las=1)


legend("topleft", legend=c("Stocks", "Cryptos"), 
       col=c("red", "black"), lty=1, cex=0.8, bty="n")

```
```{r pressure, echo=FALSE}
b1<-boxplot(combined_fd, method = "MBD")

b1<-boxplot(combined_fd, method = "BD2")

```


```{r pressure, echo=FALSE}

start_date <- min(as.Date(data$date))
data$date_numeric <- as.numeric(as.Date(data$date) - start_date)


eval_grid <- seq(min(data$date_numeric), max(data$date_numeric), length.out = 100)

basis_smooth_list <- list()  # To store basis and smooth results

for (asset in names(assets_nbasis)) {
  nbasis_val <- assets_nbasis[asset]
  
  
  basis <- create.bspline.basis(rangeval = range(data$date_numeric), nbasis = nbasis_val)
  
  
  smooth_result <- smooth.basis(data$date_numeric, data[[paste0(asset, "_scaled")]], basis)
  
  
  basis_smooth_list[[paste0(asset, "_smooth")]] <- smooth_result
}


stock_fd_list <- lapply(stocks, function(asset) basis_smooth_list[[paste0(asset, "_smooth")]]$fd)
crypto_fd_list <- lapply(cryptos, function(asset) basis_smooth_list[[paste0(asset, "_smooth")]]$fd)


stock_eval_matrix <- do.call(cbind, lapply(stock_fd_list, function(fd) eval.fd(eval_grid, fd)))
crypto_eval_matrix <- do.call(cbind, lapply(crypto_fd_list, function(fd) eval.fd(eval_grid, fd)))

stock_fd_matrix <- fdata(stock_eval_matrix, argvals = eval_grid)
crypto_fd_matrix <- fdata(crypto_eval_matrix, argvals = eval_grid)

stock_depth <- depth.FM(stock_fd_matrix)$dep
crypto_depth <- depth.FM(crypto_fd_matrix)$dep

stock_depth_df <- data.frame(Asset = stocks, Depth = stock_depth)
stock_depth_df <- stock_depth_df[order(stock_depth_df$Depth, decreasing = TRUE), ]

crypto_depth_df <- data.frame(Asset = cryptos, Depth = crypto_depth)
crypto_depth_df <- crypto_depth_df[order(crypto_depth_df$Depth, decreasing = TRUE), ]


print("Most Central Stocks:")
print(head(stock_depth_df, 3))

print("Most Outlying Stocks:")
print(tail(stock_depth_df, 3))

print("Most Central Cryptos:")
print(head(crypto_depth_df, 3))

print("Most Outlying Cryptos:")
print(tail(crypto_depth_df, 3))

```


```{r pressure, echo=FALSE}

combined_depth_df <- rbind(data.frame(Asset = stocks, Depth = stock_depth, Type = "Stock"),
                            data.frame(Asset = cryptos, Depth = crypto_depth, Type = "Crypto"))

ggplot(combined_depth_df, aes(x = reorder(Asset, Depth), y = Depth, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ Type, scales = "free_y") +
  labs(title = "Depth Values of Assets by Type", x = "Assets", y = "Depth") +
  coord_flip() +
  theme_minimal()


```

# Fourier Analysis


```{r pressure, echo=FALSE}

day_range <- c(1, nrow(data))

nbasis <- 20
b_spline_basis <- create.bspline.basis(rangeval=day_range, nbasis=nbasis)

fourier_basis <- create.fourier.basis(rangeval=day_range, nbasis=12)  # Monthly basis

smooth_and_plot <- function(asset) {
  # B-spline smoothing
  b_spline_smooth <- smooth.basis(argvals=1:nrow(data), y=data[[paste0(asset, "_scaled")]], fdParobj=b_spline_basis)
  
  # Fourier smoothing
  fourier_smooth <- smooth.basis(argvals=1:nrow(data), y=data[[paste0(asset, "_scaled")]], fdParobj=fourier_basis)
  
  
  plot(1:nrow(data), data[[paste0(asset, "_scaled")]], type='l', col='lightgray', 
       main=paste(asset, ": Original Scaled and Smoothed Curves"), ylab="Scaled Price", xlab="Time")
  lines(b_spline_smooth$fd, col="black", lwd=2, lty=1)  # B-spline
  lines(fourier_smooth$fd, col="red", lwd=2, lty=2)     # Fourier
  
  
  legend("topleft", legend=c(paste(asset, "Original"), 
                              paste(asset, "B-spline"), 
                              paste(asset, "Fourier")), 
         col=c("lightgray", "black", "red"), 
         lty=c(1, 1, 2), lwd=c(1, 2, 2))
}

for (asset in c(stocks, cryptos)) {
  smooth_and_plot(asset)
}


```


```{r pressure, echo=FALSE}

data <- data %>%
  mutate(date = as.Date(date, format = "%Y/%m/%d"))  

data_long <- data %>%
  pivot_longer(cols = -date, names_to = "asset", values_to = "price") %>%
  mutate(price_scaled = scale(price))  # Scale the prices for Fourier analysis

stocks_data <- data_long %>% filter(asset %in% stocks)  
cryptos_data <- data_long %>% filter(asset %in% cryptos)  

# Convert the date column to numeric for Fourier analysis
stocks_data <- stocks_data %>% mutate(time = as.numeric(as.Date(date)))
cryptos_data <- cryptos_data %>% mutate(time = as.numeric(as.Date(date)))

# Create Fourier basis 
fourier_basis <- create.fourier.basis(rangeval = c(min(stocks_data$time), max(stocks_data$time)), nbasis = 12)  

# Function to perform Fourier analysis on the scaled prices
fourier_analysis_group <- function(group_data) {
  # Smooth the data with the Fourier basis
  fourier_smooth <- smooth.basis(argvals = group_data$time, y = group_data$price_scaled, fdParobj = fourier_basis)
  
  # Return the Fourier coefficients
  return(fourier_smooth$fd$coefs)
}

# Perform Fourier analysis for stocks and cryptos
stocks_fourier_coeffs <- fourier_analysis_group(stocks_data)
cryptos_fourier_coeffs <- fourier_analysis_group(cryptos_data)

# Combine the coefficients into a single data frame for plotting
stocks_coeffs_df <- as.data.frame(stocks_fourier_coeffs)
cryptos_coeffs_df <- as.data.frame(cryptos_fourier_coeffs)

# Create a data frame with frequency values
frequencies <- 1:nrow(stocks_coeffs_df)
stocks_coeffs_df <- cbind(frequencies, stocks_coeffs_df)
cryptos_coeffs_df <- cbind(frequencies, cryptos_coeffs_df)

# Convert to long format for ggplot
stocks_long <- pivot_longer(stocks_coeffs_df, -frequencies, names_to = "Asset", values_to = "Coefficient")
cryptos_long <- pivot_longer(cryptos_coeffs_df, -frequencies, names_to = "Asset", values_to = "Coefficient")

# Plotting Fourier coefficients for stocks
ggplot(stocks_long, aes(x = frequencies, y = Coefficient, color = Asset)) +
  geom_line() +
  labs(title = "Fourier Coefficients: Stocks", x = "Frequency", y = "Coefficient Value") +
  theme_minimal()

# Plotting Fourier coefficients for cryptos
ggplot(cryptos_long, aes(x = frequencies, y = Coefficient, color = Asset)) +
  geom_line() +
  labs(title = "Fourier Coefficients: Cryptos", x = "Frequency", y = "Coefficient Value") +
  theme_minimal()

```





































